Verification of Stencil Computations through Spatial Specifications

### 2016-07-06 paper.tex
Revised checking, consistency, and worked example

\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{stmaryrd}
\usepackage{color}
\usepackage{graphics}
\usepackage{fancyvrb}
\usepackage{amsthm}
\usepackage{multirow}
\usepackage{minted}
\usepackage{breqn}
%\usepackage{subcaption}
\usepackage{subfigure}
\usepackage{siunitx} % For pretty-printing numeric values and SI units
                     % of measure. e.g., the tabular column type S is
                     % used to print nice-looking tables of numbers.
\sisetup{ % defaults
  group-separator={,},
  group-minimum-digits={3},
  output-decimal-marker={.},
  table-format = 6
}

@@

\begin{figure}
\vspace{-0.2em}
\begin{align*}
\setlength{\arraycolsep}{0.05em}
\begin{array}{c}
%\framebox{$\textit{region} \equiv \textit{region}'$} \\[0.7em]
\begin{array}{rll}
(\textsc{F\, \texttt{+} \,F}) \;\; &
\stenFwdSR{n \, \textsf{max} \, m}{d}{r \vee s} \\
 \equiv \; & \stenFwdSR{n}{d}{r} \; \texttt{+} \; \stenFwdSR{m}{d}{s} \\[0.7em]
%
%
(\textsc{B\, \texttt{+} \,B}) \;\; &
\stenBwdSR{n \, \textsf{max} \, m}{d}{r \vee s} \\
 \equiv \; & \stenBwdSR{n}{d}{r} \; \texttt{+} \; \stenBwdSR{m}{d}{s} \\[0.7em]
%
%
(\textsc{C\, \texttt{+} \,C}) \;\; &
\stenCenSR{n \, \textsf{max} \, m}{d}{r \vee s} \\
\equiv \; & \stenCenSR{n}{d}{r} \; \texttt{+} \; \stenCenSR{m}{d}{s} \\[0.7em]
%
%
(\textsc{C\, \texttt{+} \,F}) \;\; & \stenCenSR{n}{d}{r \vee s} \\
\textit{$m \leq n$} \; \equiv \; & \stenCenSR{n}{d}{r} \; \texttt{+} \;
                      \stenFwdSR{m}{d}{s} \\[0.7em]
%
%
(\textsc{C\, \texttt{+} \,B}) \;\; &
\stenCenSR{n}{d}{r \vee s} \\
\textit{$m \leq n$} \; \equiv \; & \stenCenSR{n}{d}{r} \; \texttt{+} \;
                      \stenBwdSR{m}{d}{s} \\[0.7em]
%
%
(\textsc{B\, \texttt{+} \,F}) \;\; &
\stenCenSR{n}{d}{r \vee s} \\
\equiv \; & \stenFwdSR{n}{d}{r} \; \texttt{+} \; \stenBwdSR{n}{d}{s}
  \\[0.7em]
%
%
(\textsc{R\, \texttt{+} \,F}) \;\; &
\stenFwdS{n}{d} \\
\equiv \; & \stenReflSR{d} \; \texttt{+} \; \stenFwdSR{n}{d}{r} \\[0.7em]
%
%
(\textsc{R\, \texttt{+} \,B}) \;\; &
\stenBwdS{n}{d} \\
\equiv \; & \stenReflSR{d} \; \texttt{+} \; \stenBwdSR{n}{d}{r} \\[0.7em]
%
%
(\textsc{R\, \texttt{+} \,C}) \;\; &
\stenCenS{n}{d} \\
\equiv \; & \stenReflSR{d} \; \texttt{+} \; \stenCenSR{n}{d}{r} \\[0.7em]
%
%
(\textsc{irrefl}) \;\; & (R \, \term{*} \, S^\irreflS) \, \term{+} \,
                         (R^\irreflS \, \term{*} \, S) \;
\equiv \; R \, \term{*} \, S \\[0.5em]
%
%
(\texttt{+}\textsc{idem}) \;\; & S \; \texttt{+} \; S \; \equiv \; S \\[0.5em]
%
(\texttt{+}\textsc{comm}) \;\; & S \; \texttt{+} \; T \; \equiv \; T \;
                       \texttt{+} \; S \\[0.5em]
%
(\texttt{+}\textsc{assoc}) \;\; & R \; \texttt{+} \; (S \; \texttt{+} \; T) \; \equiv \; (R \;
                       \texttt{+} \; S) \; \texttt{+} \; T \\[0.5em]
(\texttt{*}\textsc{comm}) \;\; & S \; \texttt{*} \; T \; \equiv \; T \;
                       \texttt{*} \; S \\[0.5em]
%
(\texttt{*}\textsc{assoc}) \;\; & R \; \texttt{*} \; (S \; \texttt{*} \; T) \; \equiv \; (R \;
                       \texttt{*} \; S) \; \texttt{*} \; T \\[0.5em]
(\textsc{dist}) \;\; & R \; \texttt{*} \; (S \; \texttt{+} \; T) \; \equiv \; (R \;
                       \texttt{*} \; S) \; \texttt{+} \; (R
                       \; \texttt{*} \; T)
\end{array}
%
%\framebox{${\textit{specDec} :: \texttt{v}} \equiv
%{\textit{specDec'} :: \texttt{v}}$} \\[0.7em]
%\begin{array}{rl}
% MUTUAL
%(\textsc{mutual}) \; &
%\texttt{stencil} \; \texttt{dependency} (\texttt{v}), \texttt{mutual}
%  :: \texttt{w}
%\\
%\equiv \; & \texttt{stencil} \; \texttt{dependency} (\texttt{w}), \texttt{mutual} ::
%  \texttt{v} \\[0.5em]
%(\textsc{coalesce}) \; &
%\texttt{stencil} \; \texttt{dependency} (\bar{v})
%  :: \texttt{v} \\
%& ; \texttt{stencil} \; \texttt{dependency} (\bar{w})
%  :: \texttt{v}
%\\
%\equiv \; & \texttt{stencil} \; \texttt{dependency} (\bar{v}, \bar{w}) ::
%  \texttt{v} \\[0.5em]
%\end{array}
\end{array}
\end{align*}
\caption{Equations on regions, $\textit{region} \equiv \textit{region}'$}
\label{fig:equations}
\end{figure}

@@

\begin{figure}[t]
\begin{align*}
\begin{array}{c}
%%
%% INEQUATIONS ON REGION
%% 
%\framebox{$\textit{region} <: \textit{region}'$} \\[0.85em]
%
\dfrac{}{R <:^r R}\trule{\textsc{refl}} \qquad \dfrac{P <:^r R \quad R <:^r Q}{P <:^r
  Q}\trule{\textsc{trans}} 
\qquad
\dfrac{R \equiv S}{R <:^r S}\trule{\textsc{eq}} 
\\[1.5em]
\setlength{\arraycolsep}{0.1em}
\dfrac{\hspace{3em} m \leq n \qquad r \Rightarrow s \hspace{3em}}
{\begin{array}{rl}
\stenFwdSR{m}{d}{r} & <:^r \stenFwdSR{n}{d}{s}
%& \;\, ds \subseteq es \wedge n \leq m
\\
\wedge \; \stenBwdSR{m}{d}{r} & <:^r \stenBwdSR{n}{d}{s}
%& \;\, ds \subseteq es \wedge n \leq m
\\
\wedge \; \stenCenSR{m}{d}{r} & <:^r \stenCenSR{n}{d}{s}
%& \;\, ds \subseteq es \wedge n \leq m
\\
\wedge \; \stenBwdSR{m}{d}{r} & <:^r \stenCenSR{n}{d}{s} \\
\wedge \; \stenFwdSR{m}{d}{r} & <:^r \stenCenSR{n}{d}{s} \\[1em]
\end{array}}\trule{\textsc{over}} \\
\hspace{-0.5em}
\dfrac{P_1 <:^r R_1 \quad P_2 <:^r R_2}
      {P_1 \, \texttt{+} \, P_2 <:^r R_1 \, \texttt{+} \, R_2}
\trule{\textsc{cong}\texttt{+}} \;\;\;
\dfrac{P_1 <:^r R_1 \quad P_2 <:^r R_2}
      {P_1 \, \texttt{*} \, P_2 <:^r R_1 \, \texttt{*} \, R_2}
\trule{\textsc{cong}\texttt{*}} \\[1.5em]
%%
%% INEQUATIONS ON SPEC
%%
%\framebox{$\textit{spec} <: \textit{spec}'$} \\[0.8em]
%
\dfrac{}{\term{readOnce} \, R <: R}\trule{\textsc{rep}}
\quad
\dfrac{}{\term{atLeast} \, R <: \term{atMost} \,
  R}\trule{\textsc{bound}} \\[1.5em]
%%
\dfrac{P <:^r R}{\term{atLeast} \, P <: \term{atLeast} \,
  R}\trule{\textsc{cong}\texttt{L}}
\ 
\dfrac{P <:^r R}{\term{atMost} \, P <: \term{atMost} \,
  R}\trule{\textsc{cong}\texttt{M}} \\[1.5em]
\dfrac{P <:^r R}{\term{readOnce} \, P <: \term{readOnce} \,
  R}\trule{\textsc{cong}\texttt{R}} 
%%
\end{array}
\end{align*}
\caption{Inequations $<:^r$ on \textit{regions} and
$<:$ on \textit{specs}.}
\label{fig:inequations}
\end{figure}

@@

\begin{definition}[Schematic]
The partial function $\textsf{schematic}$ maps 
source-language indexing terms $(e_1, \ldots, e_n)$
into $n$-dimensional index schemes 
$\in (\domainVal)^n$, defined:
%
\begin{align*}
\mathsf{schematic}(e_1, \ldots, e_n) & = 
[\downarrow\!e_1 \ldots \downarrow\!e_n ] 
\quad \textit{iff} \; \forall i . \downarrow\!e_i \neq \bot 
\end{align*}
%
The $\downarrow$ operator is a partial\footnote{We represent
  partiality explicitly in our types with the set $A + \bot$.} function 
 $\downarrow : \textit{expr} \rightarrow (\domainVal + \bot)$
which maps each individal indexing expression to a component of
the indexing scheme, defined: 
%%
\begin{align*}
\downarrow\!e
 & =  \begin{cases}
a & e \equiv i + a \; \wedge \; \mathsf{neigh}(i + a) \\
-a & e \equiv i - a \; \wedge \; \mathsf{neigh}(i - a) \\
\infty & \textit{$\mathsf{IV}(e) = \emptyset$} \\
\bot   & \textit{otherwise}
\end{cases}
\end{align*}
%%
where $\mathsf{IV}(e)$ is the set of free induction variables
 in an expression and $\neigh{}$ classifies expressions which 
are constant offsets on an induction variable
(Defn.~\ref{def:neighbour}). In the third case, 
if the expression is not a neighbourhood index but is not
a relative index either (\ie{} is not defined in terms
of an induction variable) then it is mapped to $\infty$, otherwise
$\downarrow\!e$ is undefined. The $\mathsf{schematic}$ function
is undefined if $\downarrow$ of any of its components $e_i$ is
undefined.

For example, assuming
$i$ and $j$ are induction variables and $\textsf{schematic}(i-1, 2, j)
 = \vthreeh{-1}{\infty}{0}$. However, $\textsf{schematic}(i*2, j)$, for
 example, is
 undefined as its first index is a non-neighbour relative index, which 
 is outside the scope of our specification language.
\end{definition}

@@

\begin{example}
Three array indexing terms
$(i, j+1)$, $(1, j+1)$ and $(3, j)$
are mapped by \textsf{schematic} to the 
index schemes: $\vtwoh{0}{1}$, $\vtwoh{\infty}{1}$
and $\vtwoh{\infty}{0}$ respectively. The following
relationship of generalises then holds:
%\[
%\xymatrix@R=0.2em@C=2em{
%[\infty \;\; 0] \ar@{-}[dr]^{\preceq_n} &  \\
%& [0 \;\; 1]  \\
%[\infty \;\; 1]  \ar@{-}[ur]_{\preceq_n} & 
%}
%\]
%%
\begin{equation*}
\vtwoh{\infty}{0} \; \not\preceq_2 \; \vtwoh{0}{1} 
\qquad
\vtwoh{\infty}{1} \; \preceq_2 \; \vtwoh{0}{1} 
\end{equation*}
%%
\end{example}
\noindent
Our terminology of \emph{indexing schemes} and
\emph{generalisation} is by analogy
with the \emph{type schemes} and \emph{type generalisation} of
polymorphic types in ML~\cite{milner1978theory}. Our notion is
much more rigid however, as our schemes do not bind names
and thus generalisation does not require substitution; $\infty$ acts
as a positional placeholder.

@@

\begin{example}
For a $2$-dimensional situation, 
\begin{align*}
\interp{\stenCenS{2}{1}}_2 = \; & \{i\singleEntry{1}{2} \,|\, i \in \{\text{-}2,..., 2\}\!\setminus\!0 \}
  \cup  \{ \zeroEntry{1}{2} \,|\, \textit{true} \} \\
= \; & \{-2\singleEntry{1}{2}, \, -\singleEntry{1}{2}, \,
  \singleEntry{1}{2}, \, 2\singleEntry{1}{2}\} \cup
  \{\zeroEntry{1}{2}\} \\
= \; & \{\vtwoh{-2}{\infty}, \vtwoh{-1}{\infty}, \vtwoh{0}{\infty}, 
\vtwoh{1}{\infty}, \vtwoh{2}{\infty}\} 
\end{align*}
Note the absboring behaviour of $\infty$ with respect to
multiplication. 
\end{example}
\noindent
The model of $\term{+}$ is the straightforward union of 
the models of subterms. The model of $\term{*}$ is
the tensor $\otimes$ of subterm models. The tensor 
is non-trivial; we give some informal intutition first.

@@

\begin{definition} The \textsf{Mult}
and \textsf{Approx} are parametric labelled variant types 
with injection functions given by their definition:
%
\begin{align*}
\textsf{Mult} \, a & = \textsf{mult} \, a \,\mid\, \textsf{only} \, a \\
\textsf{Approx} \, a & = \textsf{exact} \, a \,\mid\, \textsf{lower} \, a \,\mid\,
\textsf{upper} \, a \,\mid\, \textsf{both} \, a \, a
\end{align*}
\ie{}, $\mathsf{lower}$ is an injection $\mathsf{lower} : a \rightarrow \mathsf{Approx} \, a$, and so
on. The $\textsf{Mult}$ type corresponds to the presence or absence of the
\term{readOnce} modifier as shown in Figure~\ref{fig:spatial-model}. 
The $\textsf{Approx}$ type corresponds to the presence or absence of
the spatial approximation modifier, with \textsf{exact} when there is
no such modifier and \textsf{lower} and \textsf{upper} for
\term{atLeast} and \term{atMost}. Later, it will be useful to model
pairs of lower and upper bounds which is provided by \textsf{both} 
(but which isn't used here).

These are used for labelled models, and other components later in
Section~\ref{sec:analysis} with additional information.
\label{def:mult-and-approx}
\end{definition}
%

\begin{figure}[t]
\input{spatial-model.tex}
\caption{\textit{spec} model, 
$\interp{-}_n : \textit{spec} \rightarrow \textsf{Mult}(\textsf{Approx}(\mathcal{P}(\domainVal{}^n)))$}
\label{fig:spatial-model}
\vspace{-1em}
\end{figure}

@@

Given the model defined in this section we can
consider soundness for the equational theory of
specifications and the notion of spatial approximation. The proofs are given in our supplement.

\begin{theorem}[Soundness of equational theory (Figure~\ref{fig:equations})]
\[
S \equiv T \; \Rightarrow \;
\interp{S}_n = \interp{T}_n
\]
\end{theorem}

\begin{theorem}[Soundness of inequations
  (Figure~\ref{fig:inequations})]

Let $\leq$ be an ordering on
$\textsf{Mult}(\textsf{Approx}(\mathcal{P}(\domainVal{}^n)))$
and $\leq'$ be an ordering on
$\textsf{Approx}(\mathcal{P}(\domainVal{}^n))$ defined by the
following caluses
\begin{align*}
(M \leq' N \Rightarrow \textit{inj} \, M \leq \textit{inj} \, N) &
\qquad 
(\textsf{once} \, M \leq' \textsf{mult} \, M) \\
%%
%%
(M \subseteq N \Rightarrow \textit{inj}' M \leq' \textit{inj}' N)
& \;\;\; (\textsf{exact} M \leq  \textsf{lower} M)
\;\;\; (\textsf{lower} M \leq  \textsf{upper} M)
\end{align*}
%%
where $\textit{inj}$ ranges over the unary injections
of $\textsf{Approx}$ and $\textit{inj}'$ ranges over the injections of
$\textsf{Mult}$. Then, soundness of $<:$ is that:
%%
\[
S <: T
\; \Rightarrow \;
\interp{S}_n \leq \interp{T}_n
\]
\end{theorem}

@@

\noindent
We briefly sketch the algorithms for inferring and checking
specifications and for synthesising programs from specifications.
These build on the model of Section~\ref{sec:semantics}.

\paragraph{Syntax helper functions}
Let \rhsExp{} map source language statements to a
set of expressions which occur in right-hand positions (\ie{}, not the
target of an assignment). Let \var{} map expressions to a
set of the variables used in right-hand positions.

@@

\newcommand{\bplus}{\operatornamewithlimits{\term{\huge{+}}}}
\newcommand{\tySum}[1]{#1^{\term{+}}}
\newcommand{\tyProd}[1]{#1^{\term{*}}}
\newcommand{\specDNF}{\textbf{spec}}

Next, the set of spans $\spanOp(M)$ covering the indexing
space is converted into abstract syntax trees 
of the specification language. This conversion is defined in terms
of an algebra on \textit{spec} syntax in disjunctive-normal form
(which we denote \specDNF{}), and which is wrapped
with the $\textsf{Approx}$ type (see
Definition~\ref{def:mult-and-approx},
p.~\pageref{def:mult-and-approx}), which we write as $\mathsf{A}$
for brevity.  The algebra mirrors the shape
of the region combinators for addition and multiplication, 
with operators $\oplus$ and $\odot$ and a constructor for \specDNF{}:
%%
\begin{align*}
\oplus & : \mathsf{A}(\specDNF) \rightarrow \mathsf{A}(\specDNF) \rightarrow
  \mathsf{A}(\specDNF) \\
\odot & : \mathsf{A}(\specDNF) \rightarrow \mathsf{A}(\specDNF) \rightarrow
        \mathsf{A}(\specDNF) \\
\mathsf{toSpec} & : (\mathbb{Z} \cup \{\infty\}) \rightarrow (\mathbb{Z} \cup
  \{\infty\})  \rightarrow \mathsf{A}(\specDNF)
\end{align*}
%%
The $\mathsf{toSpec}$ operation takes a pair of lower and upper bounds
values (one-dimensional) and constructs a specification 
which contains a one-dimensional region. 
An entire span is converted into an $n$-dimensional
specification by applying $\mathsf{toSpec}$ to each
pair of lower and upper bounds in a span, per dimension, and multiplying these
specifications by $\odot$. The resulting specifications for
each span are then summed by $\oplus$:
\begin{equation*}
\textsf{spec}(M) = 
\sum\limits_{\vect{x} \in \spanOp(M)}
\prod\limits_{i \in \{1, \ldots, n\}} \, \mathsf{toSpec}(\vect{x}^L_i, \vect{x}^U_i)
\end{equation*}
(where $\oplus$ is the additive and $\odot$ is the multiplicative used
within $\Sigma$ and $\Pi$ respectively here).
We unpack the definitions of this algebra.

>>>>

\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{stmaryrd}
\usepackage{color}
\usepackage{graphics}
\usepackage{fancyvrb}
\usepackage{subfigure}
\usepackage{amsthm}
\usepackage{multirow}
\usepackage{minted}
\usepackage{breqn}
%\usepackage{subcaption}
\usepackage{subfigure}
\usepackage{siunitx} % For pretty-printing numeric values and SI units
                     % of measure. e.g., the tabular column type S is
                     % used to print nice-looking tables of numbers.
\sisetup{ % defaults
  group-separator={,},
  group-minimum-digits={3},
  output-decimal-marker={.},
  table-format = 6
}

@@

\begin{figure}
\vspace{-0.2em}
\begin{align*}
\setlength{\arraycolsep}{0.05em}
\begin{array}{c}
\framebox{$\textit{region} \equiv \textit{region}'$} \\[0.7em]
\begin{array}{rll}
(\textsc{F\, \texttt{+} \,F}) \;\; &
\stenFwdSR{n \, \textsf{max} \, m}{d}{r \vee s} \\
 \equiv \; & \stenFwdSR{n}{d}{r} \; \texttt{+} \; \stenFwdSR{m}{d}{s} \\[0.7em]
%
%
(\textsc{B\, \texttt{+} \,B}) \;\; &
\stenBwdSR{n \, \textsf{max} \, m}{d}{r \vee s} \\
 \equiv \; & \stenBwdSR{n}{d}{r} \; \texttt{+} \; \stenBwdSR{m}{d}{s} \\[0.7em]
%
%
(\textsc{C\, \texttt{+} \,C}) \;\; &
\stenCenSR{n \, \textsf{max} \, m}{d}{r \vee s} \\
\equiv \; & \stenCenSR{n}{d}{r} \; \texttt{+} \; \stenCenSR{m}{d}{s} \\[0.7em]
%
%
(\textsc{C\, \texttt{+} \,F}) \;\; & \stenCenSR{n}{d}{r \vee s} \\
\textit{$m \leq n$} \; \equiv \; & \stenCenSR{n}{d}{r} \; \texttt{+} \;
                      \stenFwdSR{m}{d}{s} \\[0.7em]
%
%
(\textsc{C\, \texttt{+} \,B}) \;\; &
\stenCenSR{n}{d}{r \vee s} \\
\textit{$m \leq n$} \; \equiv \; & \stenCenSR{n}{d}{r} \; \texttt{+} \;
                      \stenBwdSR{m}{d}{s} \\[0.7em]
%
%
(\textsc{B\, \texttt{+} \,F}) \;\; &
\stenCenSR{n}{d}{r \vee s} \\
\equiv \; & \stenFwdSR{n}{d}{r} \; \texttt{+} \; \stenBwdSR{n}{d}{s}
  \\[0.7em]
%
%
(\textsc{R\, \texttt{+} \,F}) \;\; &
\stenFwdS{n}{d} \\
\equiv \; & \stenReflSR{d} \; \texttt{+} \; \stenFwdSR{n}{d}{r} \\[0.7em]
%
%
(\textsc{R\, \texttt{+} \,B}) \;\; &
\stenBwdS{n}{d} \\
\equiv \; & \stenReflSR{d} \; \texttt{+} \; \stenBwdSR{n}{d}{r} \\[0.7em]
%
%
(\textsc{R\, \texttt{+} \,C}) \;\; &
\stenCenS{n}{d} \\
\equiv \; & \stenReflSR{d} \; \texttt{+} \; \stenCenSR{n}{d}{r} \\[0.7em]
%
%
(\textsc{irrefl}) \;\; & (R \, \term{*} \, S^\irreflS) \, \term{+} \,
                         (R^\irreflS \, \term{*} \, S) \;
\equiv \; R \, \term{*} \, S \\[0.5em]
%
%
(\texttt{+}\textsc{idem}) \;\; & S \; \texttt{+} \; S \; \equiv \; S \\[0.5em]
%
(\texttt{+}\textsc{comm}) \;\; & S \; \texttt{+} \; T \; \equiv \; T \;
                       \texttt{+} \; S \\[0.5em]
%
(\texttt{+}\textsc{assoc}) \;\; & R \; \texttt{+} \; (S \; \texttt{+} \; T) \; \equiv \; (R \;
                       \texttt{+} \; S) \; \texttt{+} \; T \\[0.5em]
(\texttt{*}\textsc{comm}) \;\; & S \; \texttt{*} \; T \; \equiv \; T \;
                       \texttt{*} \; S \\[0.5em]
%
(\texttt{*}\textsc{assoc}) \;\; & R \; \texttt{*} \; (S \; \texttt{*} \; T) \; \equiv \; (R \;
                       \texttt{*} \; S) \; \texttt{*} \; T \\[0.5em]
(\textsc{dist}) \;\; & R \; \texttt{*} \; (S \; \texttt{+} \; T) \; \equiv \; (R \;
                       \texttt{*} \; S) \; \texttt{+} \; (R
                       \; \texttt{*} \; T)
\end{array}
%
%\framebox{${\textit{specDec} :: \texttt{v}} \equiv
%{\textit{specDec'} :: \texttt{v}}$} \\[0.7em]
%\begin{array}{rl}
% MUTUAL
%(\textsc{mutual}) \; &
%\texttt{stencil} \; \texttt{dependency} (\texttt{v}), \texttt{mutual}
%  :: \texttt{w}
%\\
%\equiv \; & \texttt{stencil} \; \texttt{dependency} (\texttt{w}), \texttt{mutual} ::
%  \texttt{v} \\[0.5em]
%(\textsc{coalesce}) \; &
%\texttt{stencil} \; \texttt{dependency} (\bar{v})
%  :: \texttt{v} \\
%& ; \texttt{stencil} \; \texttt{dependency} (\bar{w})
%  :: \texttt{v}
%\\
%\equiv \; & \texttt{stencil} \; \texttt{dependency} (\bar{v}, \bar{w}) ::
%  \texttt{v} \\[0.5em]
%\end{array}
\end{array}
\end{align*}
\caption{Equations on specifications}
\label{fig:equations}
\end{figure}

@@

\begin{figure}[t]
\begin{align*}
\begin{array}{c}
\dfrac{}{S <: S}(\textsc{refl}) \qquad \dfrac{R <: S \quad S <: T}{R <:
  T}(\textsc{trans}) \\[1.5em]
\dfrac{}{\texttt{readOnce} \, S <: S}(\textsc{rep})
\\[1.5em]
\setlength{\arraycolsep}{0.1em}
\dfrac{\hspace{3em} m \leq n \hspace{3em}}
{\begin{array}{rl}
\stenFwdS{m}{d} & <: \stenFwdS{n}{d}
%& \;\, ds \subseteq es \wedge n \leq m
\\
\wedge \stenBwdS{m}{d} & <: \stenBwdS{n}{d}
%& \;\, ds \subseteq es \wedge n \leq m
\\
\wedge \; \stenCenS{m}{d} & <: \stenCenS{n}{d}
%& \;\, ds \subseteq es \wedge n \leq m
\\
\wedge \; \stenBwdS{m}{d} & <: \stenCenS{n}{d} \\
\wedge \; \stenFwdS{m}{d} & <: \stenCenS{n}{d} \\[1em]
\end{array}} \\[2.5em]
\hspace{-0.5em}
\dfrac{S_1 <: T_1 \quad S_2 <: T_2}
      {S_1 \, \texttt{+} \, S_2 <: T_1 \, \texttt{+} \, T_2}
(\textsc{cong}\texttt{+}) \;\;\;
\dfrac{S_1 <: T_1 \quad S_2 <: T_2}
      {S_1 \, \texttt{*} \, S_2 <: T_1 \, \texttt{*} \, T_2}
(\textsc{cong}\texttt{*})
\end{array}
\end{align*}
\caption{Inequations on regions}
\label{fig:inequations}
\end{figure}

@@

\begin{definition}[Schematic]
The partial function $\textsf{schematic}$ maps
source-language indexing terms $(e_1, \ldots, e_n)$
into $n$-dimensional index schemes
$\in (\domainVal)^n$, defined:
%
\begin{align*}
\mathsf{schematic}(e_1, \ldots, e_n) & =
[\downarrow\!e_1 \ldots \downarrow\!e_n ]
\quad \textit{iff} \; \forall i . \downarrow\!e_i \neq \bot
\end{align*}
%
The $\downarrow$ operator is a partial\footnote{We represent
  partiality explicitly in our types with the set $A + \bot$.} function
 $\downarrow : \textit{expr} \rightarrow (\domainVal + \bot)$
which maps each individal indexing expression to a component of
the indexing scheme, defined:
%%
\begin{align*}
\downarrow\!e
 & =  \begin{cases}
a & e \equiv i + a \; \wedge \; \mathsf{neigh}(i + a) \\
-a & e \equiv i - a \; \wedge \; \mathsf{neigh}(i - a) \\
\infty & \textit{$\mathsf{IV}(e) = \emptyset$} \\
\bot   & \textit{otherwise}
\end{cases}
\end{align*}
%%
where $\mathsf{IV}(e)$ is the set of free induction variables
appearing in an expression, and $\neigh{}$ classifies expressions which
are constant offsets from an induction variable
(Defn.~\ref{def:neighbour}). In the third case,
if the expression is not a neighbourhood index but is not
a relative index either (\ie{} is not defined in terms
of an induction variable) then it is mapped to $\infty$, otherwise
$\downarrow\!e$ is undefined. The $\mathsf{schematic}$ function
is undefined if $\downarrow$ of any of its components $e_i$ is
undefined.

For example $\textsf{schematic}(i, j+1) = \vtwoh{0}{1}$ assuming
$i$ and $j$ are induction variables and $\textsf{schematic}(i-1, 2, j)
 = \vthreeh{-1}{\infty}{0}$. However, $\textsf{schematic}(i*2, j)$, for
 example, is
 undefined as its first index is a non-neighbour relative index, which
 is outside the scope of our specification language.
\end{definition}

@@

\begin{example}
Take three array indexing terms
$(i, j+1)$, $(1, j+1)$ and $(3, j)$.
As given by \textsf{schematic}, these have the
corresponding schemes: $\vtwoh{0}{1}$, $\vtwoh{\infty}{1}$
and $\vtwoh{\infty}{0}$ respectively. The following
relationship of generalises then holds:
%\[
%\xymatrix@R=0.2em@C=2em{
%[\infty \;\; 0] \ar@{-}[dr]^{\preceq_n} &  \\
%& [0 \;\; 1]  \\
%[\infty \;\; 1]  \ar@{-}[ur]_{\preceq_n} &
%}
%\]
%%
\begin{equation*}
\vtwoh{\infty}{0} \; \not\preceq_2 \; \vtwoh{0}{1}
\qquad
\vtwoh{\infty}{1} \; \preceq_2 \; \vtwoh{0}{1}
\end{equation*}
%%
\end{example}
\noindent
Our terminology of \emph{indexing schemes} and
\emph{generalisation} is by analogy
with the \emph{type schemes} and \emph{type generalisation} of
polymorphic types in ML~\cite{milner1978theory}. Our notion is
much more rigid however, as our schemes do not bind names
and thus generalisation does not require substitution; $\infty$ acts
as a positional placeholder.

@@

\begin{example}
For a $2$-dimensional situation,
\begin{align*}
& \interp{\stenCenS{2}{1}}_2 \\
= \; & \{i\singleEntry{1}{2} \,|\, i \in \{\text{-}2,..., 2\}\!\setminus\!0 \}
  \cup  \{ \zeroEntry{1}{2} \,|\, \textit{true} \} \\
= \; & \{-2\singleEntry{1}{2}, \, -\singleEntry{1}{2}, \,
  \singleEntry{1}{2}, \, 2\singleEntry{1}{2}\} \cup
  \{\zeroEntry{1}{2}\} \\
= \; & \{\vtwoh{-2}{\infty}, \vtwoh{-1}{\infty}, \vtwoh{0}{\infty},
\vtwoh{1}{\infty}, \vtwoh{2}{\infty}\}
\end{align*}
Note the absboring behaviour of $\infty$ with respect to
multiplication.
\end{example}
\noindent
The model of $\term{+}$ is the straightforward union of
the models of subterms. The model of $\term{*}$ is
the tensor $\otimes$ of subterm models. The tensor
is non-trivial; we give some informal intutition first.

@@

\begin{definition} The \textsf{Mult}
and \textsf{Approx} are parametric labelled variant (sum) types
with injection function given by their definition:
%
\begin{align*}
\textsf{Mult} \, a & = \textsf{mult} \, a \,\mid\, \textsf{only} \, a \\
\textsf{Approx} \, a & = \textsf{exact} \, a \,\mid\, \textsf{lower} \, a \,\mid\,
\textsf{upper} \, a \,\mid\, \textsf{both} \, a \, a
\end{align*}
\ie{}, $\mathsf{lower}$ is an injection $\mathsf{lower} : a \rightarrow \mathsf{Approx} \, a$, and so
on. These are used for labelled models, and other components later in
Section~\ref{sec:analysis} with additional information.
\label{def:mult-and-approx}
\end{definition}
%

\begin{figure}
\input{spatial-model.tex}
\caption{\textit{spec} model,
$\interp{-}_n : \textit{spec} \rightarrow \textsf{Mult}(\textsf{Approx}(\mathcal{P}(\domainVal{}^n)))$}
\label{fig:spatial-model}
\end{figure}

@@

Given the model defined in this section we can
consider soundness for the equational theory of
specifications and the notion of spatial approximation
on specifications. The proofs are given in our supplement.

\begin{theorem}[Soundness of equational theory]
\[
\overline{\texttt{v} : S}\equiv \overline{\texttt{u} : T}
\; \Rightarrow \;
\interp{\overline{\texttt{v} : S}} = \interp{\overline{\texttt{u} : T}}
\]
\end{theorem}

\begin{theorem}[Soundness of approximations]
\[
\overline{\texttt{v} : S} <: \overline{\texttt{u} : T}
\; \Rightarrow \;
\interp{\overline{\texttt{v} : S}} \subseteq \interp{\overline{\texttt{u} : T}}
\]
\end{theorem}

@@

\noindent
We briefly sketch the algorithms for inferring and checking
specifications and for synthesising programs from specifications.
Checking and synthesis use the set-based semantic model of
Section~\ref{sec:semantics}.

\paragraph{Syntax functions}
The function \rhsExp{} maps statements to a
set of expressions which occur in right-hand positions (\ie{}, not the
target of an assignment). The function \var{} maps expressions to a
set of the variables used in right-hand positions.

@@

\newcommand{\bplus}{\operatornamewithlimits{\term{+}}}
\newcommand{\tySum}[1]{#1^{\term{+}}}
\newcommand{\tyProd}[1]{#1^{\term{*}}}
\newcommand{\specDNF}{\textbf{spec}}
\newcommand{\result}[1]{\lfloor{#1}\rceil}

Next, the set of spans $\spanOp(M)$ covering the indexing
space is converted into abstract syntax trees
for the specification language. This conversion is defined in terms
of an algebra on \textit{spec} syntax in disjunctive-normal form
(which we denote \specDNF{}).

\begin{align*}
\result{\specDNF{}} = \specDNF + ((\specDNF{} + 1) \times
  (\specDNF{} + 1))
\end{align*}

\begin{align*}
\result{\specDNF{}} & ::= \textsf{exact} \, \specDNF{}
\mid \textsf{bounds} ([\specDNF{}], [\specDNF{}]) \\
\end{align*}

The algebra mirrors the shape
of the region combinators for addition and multiplication,
with operators $\oplus$ and $\odot$ and a constructor for \specDNF{}:
%%
\begin{align*}
\oplus & : \specDNF \rightarrow \specDNF \rightarrow
  \specDNF \\
\odot & : \specDNF \rightarrow \specDNF \rightarrow
        \specDNF \\
\mathsf{toSpec} & : (\mathbb{Z} \cup \{\infty\}) \rightarrow (\mathbb{Z} \cup
  \{\infty\})  \rightarrow \specDNF
\end{align*}
%%
The $\mathsf{toSpec}$ operation takes a pair of lower and upper bounds
values (one-dimensional) and constructs a specification
which contains a one-dimensional region.
An entire span is converted into an $n$-dimensional
specification by applying $\mathsf{toSpec}$ to each
pair of lower and upper bounds in a span, per dimension, and multiplying these
specifications by $\odot$. The resulting specifications for
each span are then summed by $\oplus$:
\begin{equation*}
\textsf{spec}(M) =
\sum\limits_{\vect{x} \in \spanOp(M)}
\prod\limits_{i \in \{1, \ldots, n\}} \, \mathsf{toSpec}(\vect{x}^L_i, \vect{x}^U_i)
\end{equation*}
(where $\oplus$ is the additive and $\odot$ is the multiplicative used
within $\Sigma$ and $\Pi$ respectively here).
We unpack the definitions of this algebra.
